---
title: "Worked example of the habitat-species network concept"
author: "I. Bartomeus"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Habitat-species networks

We use data reported in Hill and Bartomeus (2016) to show the posibilities of the habitat-species network framework to analyze landscape level data.

In this study we surveyed all main habitats in 10 study sites (2 km2). The habitat types surveyed in this study were:  
a)  Corridors  
b)	Forest  
c)	Forest/grassland boundary  
d)	Non-flowering crop boundary  
e)	Semi-natural grassland  
f)	Maintained roadside  
g)	Maintained drain.  

Within each habitat types the most flowery section was surveyed along a 50m long by 3m wide transect in which bumblebee abundance, by species, was recorded for 15 minutes. Each study plot was surveyed twice between 9th July 2014 and 25th August 2014.  

First let's load and massage the data:

```{r load_data, echo=TRUE, message=FALSE, warning=FALSE, results='hide'}
#load libraries
library(reshape)
library(vegan)
library(bipartite)

#then we read and understand the data
d <- read.csv("data/powerlines3.csv", h = TRUE)
head(d)
str(d)

#And check for mistakes.
levels(d$Gen_sp) 
#We remove unidentified species.
d <- subset(d, Gen_sp != "Bombus_spp")
levels(d$Site)
#issues with swedish letters, I rename them
levels(d$Site)[1] <- "Angeby"
levels(d$Site)[4] <- "Gavsta"
levels(d$Site)[7] <- "Kottgrind"
levels(d$Site)[8] <- "Laby_Norra"
levels(d$Site)[9] <- "Laby_Sodra"

#we decide to merge both sampling rounds for now.
str(d)
d2 <- cast(d, formula = Site + Plot + Habitat + Gen_sp  ~ . , fun.aggregate = length)
head(d2)
#rename using same nomenclature as in the paper
colnames(d2) <- c("Site","Patch","Habitat","Gen_sp","Abundance" )
```

Unfortunatelly we don't have coordinates for each habitat patch, but we can still visualizing one of the networks, even if it's not spatially explicit.

##Visualization

We select one network to work with: 

```{r prep_vis, message=FALSE, warning=FALSE, results='hide'}
#subset one site
site1 <- subset(d2, Site == "Angeby")
site1
#create the network in matrix format
ntw1 <- cast(site1, Patch ~ Gen_sp, fun.aggregate = "sum", value = "Abundance")
#we can remove first column with rownames
ntw1_ <- ntw1[,-1]
ntw1 <- as.matrix(ntw1_)
colnames(ntw1) <- gsub(pattern = "Bombus_", replacement = "B. ", 
                       x = colnames(ntw1_), ignore.case = T)
#set a patch dataframe to store its propierties.
patch <- unique(site1[, c("Patch", "Habitat")])
patch$color <- c("gold", "green", "gold", "darkgreen",
                                     "grey", "green", "darkgreen",
                                     "darkgreen", "lightgreen")
rownames(ntw1) <- patch$Patch
#set also a vector of bees
bees <- cast(site1, Gen_sp ~ ., fun.aggregate = sum, value = "Abundance")
colnames(bees)[2] <- "abundance"
bees$labs <- colnames(ntw1)
```

We can use package _bipartite_ to plot it in two different ways:

```{r vis_bipartite, echo=TRUE, message=FALSE, warning=FALSE}
#prepare legend
legend <- unique(patch[,c("Habitat", "color")])
par(xpd = T) #allow plotting outside the plot
plotweb(ntw1, col.low = as.character(patch$color)) 
legend(x=0, y=0.25, as.character(legend$Habitat), pch=21,
       col="#777777", pt.bg=as.character(legend$color), 
       pt.cex=1, cex=.6, bty="n", ncol=2)
visweb(ntw1, prednames = T, preynames = T, labsize = 0.6)
```

Just inspecting the network we can see _Bombus pascuorum_ is abundant (large box) and highly connected (6 links), specially to seminatural-habitats (in green). Let's explore other visualization options. Package _igraph_ is pretty cool too:

```{r vis_igraph, message=FALSE, warning=FALSE}
#or with pretier tools:
library(igraph)
#prepare the data for igraph
links <- site1[,c("Gen_sp", "Patch", "Abundance")]
colnames(links)[3] <- "weight"
node1 <-  unique(site1[,c("Patch", "Habitat")])
colnames(node1) <- c("node", "attribute")
node1$type <- "habitat"
node2 <-  data.frame(node = unique(site1[,c("Gen_sp")]), 
                     attribute = NA,
                     type = "species")
nodes <- rbind(node1, node2)
#create igraph object
net <- graph_from_data_frame(d=links,
                             vertices=nodes, directed=F) 
# Generate colors based habitat: 
clrs <- data.frame(nod = V(net)$attribute,
                   cols = c(patch$color, rep("blue", 14)))
V(net)$color <- as.character(clrs$cols)
# Compute node degrees (#links) and use that to set node size:
deg <- degree(net, mode="all")
V(net)$size <- deg*3
# Setting them to NA will render no labels:
V(net)$label <- as.character(nodes$id)
# Set edge width based on weight:
E(net)$width <- E(net)$weight/3
#change arrow size and edge color:
E(net)$arrow.size <- .2 #but note no arrows in Unidirected graphs like this
E(net)$edge.color <- "gray80"
#prepare colors
cl <- unique(clrs)
cl$nod <- as.character(cl$nod)
cl$nod[which(is.na(cl$nod))] <- "Bombus"
plot(net, vertex.label = NA) #force vertex label NA to make visualization clearer.
legend(x=-1.5, y=-1.1, cl$nod, pch=21,
       col="#777777", pt.bg=as.character(cl$cols), 
       pt.cex=2, cex=.8, bty="n", ncol=2)
```

Again, we can see there are 4 bigger habitat nodes that are highly connected. Interestingly, those belong to 4 different habitats. There is one more packege visNetwork that makes impresive interactive plots that may be better to visualize and explore this kind of graphs.

```{r vis_visNetwork, message=FALSE, warning=FALSE}
library('visNetwork') 
colnames(nodes)[1] <- "id"
nodes$shape <- "dot"  
nodes$shadow <- TRUE # Nodes will drop shadow
nodes$attribute <- as.character(nodes$attribute) 
nodes$attribute[10:23] <-  as.character(nodes$id[10:23])
nodes$title <- nodes$attribute # Text on click
nodes$label <- nodes$type # Node label
nodes$size <- deg*3 # Node size
nodes$borderWidth <- 2 # Node border width

nodes$color.background <- clrs$cols
nodes$color.border <- "black"
nodes$color.highlight.background <- "orange"
nodes$color.highlight.border <- "darkred"
links$width <- links$weight # line width
links$color <- "gray"    # line color  
#links$arrows <- "middle" # arrows: 'from', 'to', or 'middle'
links$smooth <- TRUE    # should the edges be curved?
links$shadow <- FALSE    # edge shadow
colnames(links)[1:2] <- c("from", "to")
visNetwork(nodes, links) 
```

Next let's calculate a few network parameters. 

##Network structure

How is this network structured? To calculate it's _nestedness_, we use weighted NODF, one of the more popular and robust nestedness metrics (Almeida-Neto et al. 2008, Almeida-Neto et al. 2010), but other options are available.

```{r nestedness, message=FALSE, warning=FALSE}
(obs <- networklevel(web = ntw1, index = "weighted NODF"))
```

To know if `r obs` is more nested than expected by chance for this network, we need to compare it with a null model:

```{r null, message=FALSE, warning=FALSE}
nm <- nullmodel(web = ntw1, N=1000, method="r2d")
null <- unlist(sapply(nm, networklevel, index="weighted NODF")) 
plot(density(null), xlim=c(min(obs, min(null)), max(obs, max(null))), 
		main="comparison of observed with null model Patefield")
abline(v=obs, col="red", lwd=2)    
praw <- sum(null>obs) / length(null)
```

We can see this network is more nested than expected by chance (p value = `r ifelse(praw > 0.5, 1-praw, praw)`). This indicates that patches hosting less diversity tend to host also common species, or from the other point of view, that rare species tend to be found in more biodiverse patches. Now, let's calculate another structural metric. Its modularity. We will use a quantitative version for this end (Dormann et al 2016).

```{r mod, message=FALSE, warning=FALSE}
res <- computeModules(ntw1)
plotModuleWeb(res, displayAlabels = T)
#listModuleInformation(res)
#printoutModuleInformation(res)
```
We can identify 4 modules. To see if this modularity level is larger than expected by chance we can follow a similar approach as above, but let's focus now on the description of this modules. We can see patches 7, 8 and 9 form a dense module with common bumblebee species. Those are forested areas. Patches 1,2 and 6 (grasslands and open habitats) host rarer bumblebees. The other modules are very small, but it's interesting that the only roadside patch has it's own module.

```{r mod2, message=FALSE, warning=FALSE}
#we can calculate 2 values for each node
cz <- czvalues(res, weighted = TRUE, level = "lower")
#c : among-module connectivity
#z : within-module connectivity
#Olesen et al. (2006) give critical c and z values of 0.62 and 2.6, respectively. Species exceeding these values are deemed connectors or hubs of a network. The justification of these thresholds remains unclear to me. They may also not apply for the quantitative version.
plot(cz[[1]], cz[[2]], pch=16, xlab="c", ylab="z", 
     cex=0.8, las=1, col = patch$col)
text(cz[[1]], cz[[2]], names(cz[[1]]), pos=4, cex=0.7)
#we can check congruence between runs wth function metaComputeModules
#res2 <- metaComputeModules(as.matrix(ntw1))
```

We can also see the among-module (z) and within-module (c) connectivity of different patches. Interestingly, module 9 and 1 and 3 tend to act as links among modules. This makes snese as patch 9 is the interfase between Forest and grassland boundaries. Note c can't be computed for modules composed by a single patch, and hence patch 5 is not plotted.

Next we can ask which patches would we prioritize its conservation. For this we calculate its strength (Bascopmte et al 2006).

```{r strength, message=FALSE, warning=FALSE}
#we traspose the matrix to calculate it for the lower level.
patch$strength <- bipartite::strength(t(ntw1), type="Bascompte")
patch
```

Interesting, roadsites has the highest strenth! This is because it sustain both common and rare species. However, we may want to correct for the fact that this habitat is only represetented once in the dataset. Not surprisingly forests tend to rank lower, as they host moderate number of common species.

Let's see now how each patch influences each other (Müller et al 1999):

```{r influ, message=FALSE, warning=FALSE}
(inf <- PAC(ntw1))
```

We can read this matrix as the influnce mediated by shared pollinators between each pair of patches. Looks like influences are overall low (mean = `r mean(inf)`) but some patches influence each others via shared pollinators (e.g. 7->8 `r inf[7,8]`, but note this is not reciprocal, as 8->7 influence is moderate: `r inf[8,7]`) 

Finally let's see which are the more selective bumblebees (Blüthgen et al 2007)? For this we regroup at habitat level.

```{r regroup, message=FALSE, warning=FALSE}
d3 <- cast(d, formula = Site + Habitat + Gen_sp  ~ . , fun.aggregate = length)
colnames(d3) <- c("Site","Habitat","Gen_sp","Abundance" )
site1b <- subset(d3, Site == "Angeby")
#create the network in matrix format
ntw1b <- cast(site1b, Habitat ~ Gen_sp, fun.aggregate = "sum", value = "Abundance")
#we can remove first column with rownames
ntw1b <- ntw1b[,-1]
#let's visualize it with bipartite
#plotweb(ntw1b)
#and calculate d'
#here low.abun can be used if we know patch attributes like area.
bees$d <- specieslevel(web = ntw1b, index = "d", level = "higher")
bees
```

We can see _B. terrestris_ or _B. pascuorum_ has high values of d' (very unselective), while _B. humilis_ or _B. soroeensis_ are highly selective.

let's scale this analysis at the site levels. We can ask which site is more nested? and if the nestedness pattern correlates with amount of semi-natural habitats or species richness levels? We can also if more nested sites are also more robust (Memmot et al. 2004)?

Let's calculate nestedness for all networks as well as other network parameters like species richness, robustness and number of semi-natural patches. Remember that to compare nestedness values, we need to standardize them. Song et al 2017 method is the more robust:


```{r nested_comms, message=FALSE, warning=FALSE}
source("toolbox.R") #load code developed by Song et al. and available in his paper.
#let's loop through all sites
#first we create empty objects to store the data
sites <- unique(d2$Site)
ntwks <- list()
nested <- c()
NODF <- c()
st_NODF <- c()
rob_rand <- c()
rob_real <- c()
rich <- c()
seminat <- c()
for(i in 1:length(sites)){
  sitex <- subset(d2, Site == sites[i])
  #create the network in matrix format
  ntwx <- cast(sitex, Patch ~ Gen_sp, fun.aggregate = "sum", 
               value = "Abundance")
  #we can remove first column with rownames
  ntwx <- ntwx[,-1]
  #let's visualize it with bipartite
  #plotweb(ntwx)
  ntwks[[i]] <- ntwx
  #calculate nestedness
  nested[i] <- networklevel(web = ntwx, index = "weighted NODF")
  rob_rand[i] <- robustness(second.extinct(web = ntwx, nrep = 50, participant = "lower", method = "random"))
  NODF[i] <- nestedness_NODF(ntwx)
  st_NODF[i] <- comb_nest(web = ntwx, NODF = NODF, max_NODF = max_nest(ntwx))
  #reate a realistic extinction sequence
  ext_seq <- unique(sitex[,c("Patch", "Habitat")])
  #quick and dirty way to order habitats
  levels(ext_seq$Habitat) <- c("gCorridor", "bForest",
                               "cForest_grassland_boundary",
                               "fMaintained_drain", 
                               "eMaintained_roadside",
                               "dNon_flowering_crop_edge",
                               "aSemi_natural_grasslands")
  ext_seq$Patch <- order(as.character(ext_seq$Habitat))
  rob_real[i] <- robustness(second.extinct(web = ntwx, participant = "lower", method = "external", ext.row = ext_seq$Patch )) #garsslands first, forest, etc...
  rich[i] <- ncol(ntwx)
  seminat[i] <- length(subset(ext_seq, Habitat %in%
                                c("aSemi_natural_grasslands",
                                  "bForest"))$Patch)
}
sites_measures <- data.frame(sites, nested, NODF, st_NODF, rob_rand, rob_real, rich, seminat)
```

_Are more nested networks correlated with the amount of semi-natural habitats? _

```{r plot1, message=FALSE, warning=FALSE}
plot(sites_measures$st_NODF ~ sites_measures$seminat)
abline(lm(sites_measures$st_NODF ~ sites_measures$seminat))
```

We are not going to run the stats here, but looks like there is a weak trend for sites with more seminatural habitat patches (forests and grasslands) to be less nested. For a proper test we would need better information on the proportion of semi-natural habitats in the landscape.

_Are more nested networks also more biodiverse? _

```{r plot2, message=FALSE, warning=FALSE}
scatter.smooth(sites_measures$st_NODF ~ sites_measures$rich)
```

In this case, the plot suggest that nestedness is not realted to bumblebee richness levels.

_Are more nested sites also more robust to in silico patch removal?_

```{r plot3, message=FALSE, warning=FALSE}
scatter.smooth(sites_measures$rob_rand ~ sites_measures$rob_real)
scatter.smooth(sites_measures$st_NODF ~ sites_measures$rob_rand)
scatter.smooth(sites_measures$neste ~ sites_measures$rob_real)
```

In this case, both the random removal of patches, and the "semi-natural habitats first" removal are correlated.Contrary to expectations, in this situation robustnes is negatively correlated with nestedness level when removals are random, but this relationship is more complex when semi-natural patches are removed first.

##What we have learned?

We have seen here how to apply network tools to anayze habitat-species networks, and highlight that the type of analysis will depend on your question and the type of data available.

##References

Almeida-Neto, M., Loyola, R.D., Ulrich, W., Guimaraes, P., Guimaraes, Jr., P.R. 2008. A consistent metric for nestedness analysis in ecological systems: reconciling concept and measurement. Oikos 117, 1227–1239

Almeida-Neto, M. & Ulrich, W. (2011) A straightforward computational approach for measuring nestedness using quantitative matrices. Environmental Modelling & Software 26, 173–178

Bascompte, J., Jordano, P. and Olesen, J. M. 2006 Asymmetric coevolutionary networks facilitate biodiversity maintenance. Science 312, 431–433

Blüthgen, N., Menzel, F., Hovestadt, T., Fiala, B. and Blüthgen N. 2007 Specialization, constraints and conflicting interests in mutualistic networks. Current Biology 17, 1–6

Burgos, E., H. Ceva, R.P.J. Perazzo, M. Devoto, D. Medan, M. Zimmermann, and A. Maria Delbue (2007) Why nestedness in mutualistic networks? Journal of Theoretical Biology 249, 307–313

Dormann, C. F., and R. Strauß. 2014. Detecting modules in quantitative bipartite networks: the QuanBiMo algorithm. Methods in Ecology & Evolution 5 90–98 (or arXiv [q-bio.QM] 1304.3218.)

Memmott, J., Waser, N. M. and Price M. V. 2004 Tolerance of pollination networks to species extinctions. Proceedings of the Royal Society B 271, 2605–2611

Müller, C. B., Adriaanse, I. C. T., Belshaw, R. and Godfray, H. C. J. 1999 The structure of an aphid-parasitoid community. Journal of Animal Ecology 68, 346–370